Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-07-27 paper/main.tex
intro

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{url}
% \usepackage[breaklinks=true]{hyperref} breaks compilation?
% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

\subsection{Abstract Data Representation}
\label{sec:atct:fiber-bundles}
We model data using a mathematical representation of data that can encode topological properties, field types, and data values in a uniform manner using a structure from algebraic topology called a fiber bundle. We extend Butler's proposal of bundles as abstract visualization data type \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} by incorporating Spivak's methodology for encoding named data types from his fiber bundle representation of relational databases \cite{spivakDatabasesAreCategories2010, spivakSIMPLICIALDATABASES}. We build on this work to describe how to encode the connectivity of the data as a topological space, separately encode the fields as their own topological space with a typing system, and express the mappings between these two spaces.

@@

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, equivariance is traditionally evaluated as the preservation of field structure, which is based on the field type. To have a rich language for expressing field type, we generalize Spivak's fiber bundle formulation of table \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.

\begin{definition} A \textbf{type specification} is the bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is the set of data types, and the total space $\mathscr{U}$ is the set of domains of those data types. Given an arbitrary data type $T \in \textbf{DT}$, the preimage $\pi^{-1} \subset \mathscr{U}$ is the domain of $T$ and $x \in \pi^{-1}(T)$ is an object of type $T$.  
\end{definition}
For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers. The domain $\mathscr{U}_{{\sigma}}$ is also a topological space, meaning there is no restriction on the dimensionality of the field, thus allowing for the encoding of multi-dimensional fields in the same manner as single dimensional fields. Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a  data base schema \cite{ullmanFirstCourseDatabase2008}. 
\begin{definition} A \textbf{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type. 
\end{definition}
The function $\sigma$ is composed with $\pi$ such that the bundle  $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ associates a field name $c \in C$ with the corresponding data type domain. Spivak then defines data records as sections on the bundle defined by $\pi_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then constructs tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of records $\Gamma^{\pi}(\sigma)$ on the schema bundle. Generalizing Spivak's formulation of tables, we define the fiber space in this work as the set of record functions 
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. The construction of the fiber as the space of records provides a way to distinguish data fields with the same datatype while allowing that datatype to be defined as almost any arbitrary topological space.

@@

While the presheaf is a generalization of the concept of a global space of sections $\Gamma(\dbase, \dtotal)$ \cite{spanier1989algebraic}, the locality and gluing axioms \cite{bakerMathsSheaf} of a sheaf describe the conditions for verifying that the sections over subspaces are part of a cohesive section over the whole space. 
\begin{definition} A \textbf{sheaf} is a presheaf that satisfies the following two axioms
\begin{itemize}
  \item \textit{locality} two sections in a sheaf are equal $\dsection^{a} = \dsection^{b}$ when they evaluate to the same values over the same set of open sets $\dsection^{a}|{\openset} =  \dsection^{b}|_{\openset}$.
  \item \textit{gluing} the union of sections defined on specific open sets is equivalent to one big section over the union of spaces $\dsection|_{\openset_{i} \cup \openset_{j}} = \dsection^{i}|_{\openset_i} \cup \dsection^{j}|_{\openset_j}$ if these sections agree on overlaps $\dsection^{i}|_{\openset_i\cap\openset_j} =  \dsection^{j}|_{\openset_i\cap\openset_j}$
  \end{itemize}
\end{definition}
The gluing axiom says that a distributed representation of a dataset, which is a set of local sections, is equivalent to a section over the union of the opensets of the local sections. The locality axiom asserts that the glued section function is equivalent to a function over the union if they evaluate t the same values. The gluing axiom can also be used to generate the gluing rules used to construct non-trivial bundles from the set of trivial local section. Generally, the sheaf asserts the expectation that the data container is implemented such that the connectivity between the opensets (underlying data continuity) is preserved.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography}

>>>>

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{xurl} 
%\usepackage{hyperref} breaks compilation?
% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

\subsection{Abstract Data Representation}
\label{sec:atct:fiber-bundles}
We model data using a mathematical representation of data that can encode topological properties, field types, and data values in a uniform manner using a structure from algebraic topology called a fiber bundle. We extend Butler's proposal of bundles as abstract visualization data type \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} by incorporating Spivak's methodology for encoding named data types from his fiber bundle representation of relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. We build on this work to describe how to encode the connectivity of the data as a topological space, separately encode the fields as their own topological space with a typing system, and express the mappings between these two spaces.

@@

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, equivariance is traditionally evaluated as the preservation of field structure, which is based on the field type. To have a rich language for expressing field type, we generalize Spivak's fiber bundle formulation of table \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. 
\note{def 3.5 and 3.6 should maybe be pushed back to prose as 3.7 is the only one we rely on - it defines what tau returns}
\begin{definition} A \textbf{type specification} is the bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is the set of data types, and the total space $\mathscr{U}$ is the set of domains of those data types. Given an arbitrary data type $T \in \textbf{DT}$, the preimage $\pi^{-1} \subset \mathscr{U}$ is the domain of $T$ and $x \in \pi^{-1}(T)$ is an object of type $T$.  
\end{definition}
For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers. The domain $\mathscr{U}_{{\sigma}}$ is also a topological space, meaning there is no restriction on the dimensionality of the field, thus allowing for the encoding of multi-dimensional fields in the same manner as single dimensional fields. Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a  data base schema \cite{ullmanFirstCourseDatabase2008}. 
\begin{definition} A \textbf{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type. 
\end{definition}
The function $\sigma$ is composed with $\pi$ such that the bundle  $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ associates a field name $c \in C$ with the corresponding data type domain. Spivak then defines data records as sections on the bundle defined by $\pi_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then constructs tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of records $\Gamma^{\pi}(\sigma)$ on the schema bundle. Generalizing Spivak's formulation of tables, we define the fiber space in this work as the set of record functions 
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. The construction of the fiber as the space of records provides a way to distinguish data fields with the same datatype while allowing that datatype to be defined as almost any arbitrary topological space.

@@

While the presheaf is a generalization of the concept of a global space of sections $\Gamma(\dbase, \dtotal)$ \cite{spanier1989algebraic}, the locality and gluing axioms \cite{bakerEuclideanSpaceMathsSheaf} of a sheaf describe the conditions for verifying that the sections over subspaces are part of a cohesive section over the whole space. 
\begin{definition} A \textbf{sheaf} is a presheaf that satisfies the following two axioms
\begin{itemize}
  \item \textit{locality} two sections in a sheaf are equal $\dsection^{a} = \dsection^{b}$ when they evaluate to the same values over the same set of open sets $\dsection^{a}|{\openset} =  \dsection^{b}|_{\openset}$.
  \item \textit{gluing} the union of sections defined on specific open sets is equivalent to one big section over the union of spaces $\dsection|_{\openset_{i} \cup \openset_{j}} = \dsection^{i}|_{\openset_i} \cup \dsection^{j}|_{\openset_j}$ if these sections agree on overlaps $\dsection^{i}|_{\openset_i\cap\openset_j} =  \dsection^{j}|_{\openset_i\cap\openset_j}$
  \end{itemize}
\end{definition}
The gluing axiom says that a distributed representation of a dataset, which is a set of local sections, is equivalent to a section over the union of the opensets of the local sections. The locality axiom asserts that the glued section function is equivalent to a function over the union if they evaluate t the same values. The gluing axiom can also be used to generate the gluing rules used to construct non-trivial bundles from the set of trivial local section. Generally, the sheaf asserts the expectation that the data container is implemented such that the connectivity between the opensets (underlying data continuity) is preserved.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography}
